\vspace{-0.5in}{\bfseries\scshape Transfer Learning for computationally limited training of Vision Transformers on art classification tasks } \\ \vspace{1ex} \normalsize{Bachelor's Project Thesis}\vspace{-2ex}

### 2022-07-30 writing/tex/introduction.tex
Made changes suggested by Grammarly

\section{Introduction}
Since the introduction of AlexNet, roughly a decade ago, \textit{Convolutional Neural Networks} (CNNs) have played an important role in \textit{Computer Vision} (CV) \citep{krizhevsky2012imagened}. Recently, however, a new type of architecture, called \textit{Vision Transformer} (VT), gained state of the art performance on common learning benchmarks, including the ImageNet dataset \citep{deng2009imagenet}\footnote{At the time of writing, the VT \texttt{CoCa} is state of the art, according to \url{https://paperswithcode.com/sota/image-classification-on-imagenet}.}.

@@

CNNs are largely invariant to shifts and small distortions of the input image. More important, is that they can be seen as automatic feature extractors. It is well established (e.g. \citeauthor{lecun1998gradient}, \citeyear{lecun1998gradient}; \citeauthor{yosinski2014transferable}, \citeyear{yosinski2014transferable}) that the first layers of a CNN detect low level features, such as edges and corners. These are combined in subsequent layers to detect higher order features. The final layer should then give a representative summary of the input, such that, for example, an MLP-type layer can extract all the information it needs from it to perform the task at hand.

@@

\paragraph{A limitation of VTs}
While VTs are promising; the fact that they are not specifically designed for CV tasks, makes that they lack a so called image specific \textit{Inductive Bias}, which CNNs do have. Intuitively this means a VT first has to learn what an image is, before it can move on to the learning task at hand\footnote{Recall, for example, that the architecture ignores the 2D layout of an image, and hence has to learn about this through training.}. The result is that they require substantially more training data (and time) than CNNs do.

It is for this reason the paper that first introduced VTs \citep{dosovitskiy2020image} mentions using pre-trained models rather than randomly initialised ones, as it allows general knowledge about images to be carried over to the new task. This technique falls within the category of TL, which will be covered next.

@@

In practice, many TL techniques exist. The example given in section \ref{vits} can be described as \textit{fine-tuning} (FT). Here, a pre-trained model is used as a starting point, and then trained on the target task -- possibly with different hyperparameters, and often with a new final layer that fits the task. \citeauthor{yosinski2014transferable} (\citeyear{yosinski2014transferable}) studied transferability properties of CNNs pre-trained on ImageNet, and suggested that FT results in improved performance and generalizability compared to training from scratch. Moreover, while transferability decreased as source and target task became more dissimilar, utilizing TL still seemed to be preferred.

@@

Another TL technique is called \textit{off-the-shelf} (OTS) learning. It is similar to FT, except that all but the final layer is kept frozen during training. \citeauthor{sharif2014cnn} (\citeyear{sharif2014cnn}) found that features extracted by a CNN pre-trained on ImageNet were general enough for OTS learning to work. More specifically, these OTS CNNs showed better performance than the (fully trained non-CNN) state of the art at the time. Besides reducing computational cost of training, a benefit of OTS learning is that it can prevent overfitting in cases where FT would not \citep{yosinski2014transferable}.

@@

What \citeauthor{matsoukas2021time} did not investigate, is OTS performance of VTs. This was, however, done in \citeauthor{zhou2021convnets} (\citeyear{zhou2021convnets}), which found that VTs pre-trained on ImageNet had worse OTS performance than their CNN counterparts. An exception was the \texttt{Swin} architecture, which was the best performing model on all datasets, and uses \textit{(shifted-)window-based} self-attention methods, to make Transformers better suited for CV tasks \citep{liu2021swin}.
% Not into detail about ots and other techniques

Another `shortcomming' of \citeauthor{matsoukas2021time} is that it merely examined whether VTs can use TL to outperform CNNs in a domain that happens to be characterised by limited-sized datasets, but that it did not actively focus on the extent to which VTs can be made useful in limited data/compute environments. More related to this is \citeauthor{touvron2021training} (\citeyear{touvron2021training}), which proposed transformer-specific distillation-based learning techniques to make VTs applicable for smaller datasets. This is not related to FT or OTS learning, of course.
% matsoukas2021time not specifically focussed on seeing to what degree TL allows VTs to dominate CNNs in limited compute/data environments
\\\\
The current thesis will examine to what extent TL techniques (both FT and OTS) allow VTs to become the preferred architecture when resources are limited, and attempts to find a methodology that is recommended in these circumstances. Several experiments will be performed, of which results will be analysed both quantitatively as well as qualitatively.

Limited resources, in this case, primarily refers to the amount of available training data. Datasets should be so small that one can reasonable assume they were hand labelled by a single person. In addition, it also refers to compute resources, which should be restricted to consumer-grade hardware. In practice, of course, the small datasets already ensure that this restriction is met, because these naturally allow such hardware to suffice for learning.

Furthermore, this thesis restricts itself to art classification tasks specifically. These tasks fit in well with the described objectives, since the amount of effort needed to create and digitise an art collection often makes for small datasets. Moreover, they provide an interesting TL challenge, as knowledge might have to be carried over from natural tot non-natural images.
%The current study limits itself to the task of identifying which materials were used in artworks. For this, it will examine to what extent TL techniques (both fine-tuning and OTS) allow VTs to become the preferred architecture when compute resources are limited, and attempt to find a methodology that is preferred in circumstances as described. It will do this by means of several experiments, and both quantitative and qualitative analysis of the results.

To put all of this more formally, the following research question is proposed: \begin{quote}\textit{To what extent (and in what manners) does utilising TL allow VTs to perform on par with CNNs on common art classification problems, when datasets contain 10,000 images or less?}\end{quote}

>>>>

\section{Introduction}
Since the introduction of AlexNet, roughly a decade ago, \textit{Convolutional Neural Networks} (CNNs) have played an important role in \textit{Computer Vision} (CV) \citep{krizhevsky2012imagened}. Recently, however, a new type of architecture, called \textit{Vision Transformer} (VT), gained state-of-the-art performance on common learning benchmarks, including the ImageNet dataset \citep{deng2009imagenet}\footnote{At the time of writing, the VT \texttt{CoCa} is state of the art, according to \url{https://paperswithcode.com/sota/image-classification-on-imagenet}.}.

@@

CNNs are largely invariant to shifts and small distortions of the input image. More important, is that they can be seen as automatic feature extractors. It is well established (e.g. \citeauthor{lecun1998gradient}, \citeyear{lecun1998gradient}; \citeauthor{yosinski2014transferable}, \citeyear{yosinski2014transferable}) that the first layers of a CNN detect low-level features, such as edges and corners. These are combined in subsequent layers to detect higher-order features. The final layer should then give a representative summary of the input, such that, for example, an MLP-type layer can extract all the information it needs from it to perform the task at hand.

@@

\paragraph{A limitation of VTs}
While VTs are promising; the fact that they are not specifically designed for CV tasks makes that they lack a so-called image-specific \textit{Inductive Bias}, which CNNs do have. Intuitively this means a VT first has to learn what an image is, before it can move on to the learning task at hand\footnote{Recall, for example, that the architecture ignores the 2D layout of an image, and hence has to learn about this through training.}. The result is that they require substantially more training data (and time) than CNNs do.

It is for this reason the paper that first introduced VTs \citep{dosovitskiy2020image} mentions using pre-trained models rather than randomly initialized ones, as it allows general knowledge about images to be carried over to the new task. This technique falls within the category of TL, which will be covered next.

@@

In practice, many TL techniques exist. The example given in section \ref{vits} can be described as \textit{fine-tuning} (FT). Here, a pre-trained model is used as a starting point, and then trained on the target task -- possibly with different hyperparameters, and often with a new final layer that fits the task. \citeauthor{yosinski2014transferable} (\citeyear{yosinski2014transferable}) studied transferability properties of CNNs pre-trained on ImageNet, and suggested that FT results in improved performance and generalizability compared to training from scratch. Moreover, while transferability decreased as the source and target task became more dissimilar, utilizing TL still seemed to be preferred.

@@

Another TL technique is called \textit{off-the-shelf} (OTS) learning. It is similar to FT, except that all but the final layer is kept frozen during training. \citeauthor{sharif2014cnn} (\citeyear{sharif2014cnn}) found that features extracted by a CNN pre-trained on ImageNet were general enough for OTS learning to work. More specifically, these OTS CNNs showed better performance than the (fully trained non-CNN) state-of-the-art at the time. Besides reducing computational cost of training, a benefit of OTS learning is that it can prevent overfitting in cases where FT would not \citep{yosinski2014transferable}.

@@

What \citeauthor{matsoukas2021time} did not investigate, is the OTS performance of VTs. This was, however, done by \citeauthor{zhou2021convnets} (\citeyear{zhou2021convnets}), which found that VTs pre-trained on ImageNet had worse OTS performance than their CNN counterparts. An exception was the \texttt{Swin} architecture, which was the best performing model on all datasets, and uses \textit{(shifted-)window-based} self-attention methods, to make Transformers better suited for CV tasks \citep{liu2021swin}.
% Not into detail about ots and other techniques

Another `shortcomming' of \citeauthor{matsoukas2021time} is that it merely examined whether VTs can use TL to outperform CNNs in a domain that happens to be characterized by limited-sized datasets, but that it did not actively focus on the extent to which VTs can be made useful in limited data/compute environments. More related to this is \citeauthor{touvron2021training} (\citeyear{touvron2021training}), which proposed transformer-specific distillation-based learning techniques to make VTs applicable for smaller datasets. This is not related to FT or OTS learning, of course.
% matsoukas2021time not specifically focussed on seeing to what degree TL allows VTs to dominate CNNs in limited compute/data environments
\\\\
The current thesis will examine to what extent TL techniques (both FT and OTS) allow VTs to become the preferred architecture when resources are limited, and attempts to find a methodology that is recommended in these circumstances. Several experiments will be performed, of which results will be analyzed both quantitatively as well as qualitatively.

Limited resources, in this case, primarily refers to the amount of available training data. Datasets should be so small that one can reasonably assume they were hand labeled by a single person. In addition, it also refers to compute resources, which should be restricted to consumer-grade hardware. In practice, of course, the small datasets already ensure that this restriction is met, because these naturally allow such hardware to suffice for learning.

Furthermore, this thesis restricts itself to art classification tasks specifically. These tasks fit in well with the described objectives, since the amount of effort needed to create and digitize an art collection often makes for small datasets. Moreover, they provide an interesting TL challenge, as knowledge might have to be carried over from natural to non-natural images.
%The current study limits itself to the task of identifying which materials were used in artworks. For this, it will examine to what extent TL techniques (both fine-tuning and OTS) allow VTs to become the preferred architecture when compute resources are limited, and attempt to find a methodology that is preferred in circumstances as described. It will do this by means of several experiments, and both quantitative and qualitative analysis of the results.

To put all of this more formally, the following research question is proposed: \begin{quote}\textit{To what extent (and in what manners) does utilizing TL allow VTs to perform on par with CNNs on common art classification problems, when datasets contain 10,000 images or less?}\end{quote}

### 2022-07-30 writing/tex/methods.tex
Made changes suggested by Grammarly

% Experiments build upon Sabatelli et al.
Much of the experiments and analysis discussed below, builds upon the work done in \citeauthor{sabatelli2018deep} (\citeyear{sabatelli2018deep}). This paper investigated TL properties of CNNs in the art classification domain. The main focus was on a comparison between CNN architectures pre-trained on ImageNet1K when they were either used as OTS feature extractors, or fine-tuned to the new task. It was suggested that fine-tuning resulted in substantially better performance than either training from scratch or using an OTS learning scheme.

@@

Similar to \citeauthor{sabatelli2018deep}, used models are pre-trained on ImageNet1K, and the \textit{Rijksmuseum Challenge} dataset \citep{mensink14icmr} is used for the target tasks. This dataset consists of a large collection of digitized artworks, together with xml-formatted metadata. The classification tasks extracted from it are: (1) \textit{Type classification}, where the model has to distinguish between classes as `painting', `sculpture', `drawing', etc.; (2) \textit{Material classification}, with classes as `paper', `porcelain', `silver', etc.; and finally, (3) \textit{Artist classification}, where the model has to predict who the creator is. Figure \ref{methods:eg} shows labelled examples for all three tasks.

@@

There are still many samples and classes left after doing the operations described above. Samples are also unevenly distributed among classes, with Gini-coefficients being as high as 0.98\footnote{The Gini-coefficient is a measure of balance, with 0 meaning all classes have the same number of samples, and 1 meaning all samples belong to a single class. A uniform distribution has coefficient $\frac{1}{3}$}. To counteract this, only the 30 most occurring classes are selected, and a cap of 1000 instances per class is enforced by taking a random sample when this limit is exceeded. Table \ref{methods:datasets} shows the result of these balancing and subsampling operations. The situation prior is shown between brackets.

@@

All datasets are split up into 80\%, 10\%, 10\% training, validating and testing sets. For robustness, five datasets are randomly generated per classification task, such that each experiment can be run five times. The average sample overlap between two of the five sets is shown in table \ref{methods:datasets} as well.

Besides the tasks described above, table \ref{methods:datasets} shows an experiment called \textit{Scaling}. The goal of this experiment is to see how the findings of this paper hold up as dataset sizes are gradually shrunk. It does this by taking the top 15 classes of type classification, and scaling it 4 times by a factor $\sqrt[4]{\frac{1}{10}} \approx 0.56$. The 5 datasets produced in this manner are then respectively 100\%, 56\%, 32\%, 18\% and 10\% the size of the dataset shown in table \ref{methods:datasets}. Steps are taken to ensure that the distribution remains the same. Finally, also here 5 datasets are generated per scaling factor, leading to a total of 25 datasets. The same 80\%, 10\%, 10\% split is used for all.

@@

For all models, the linear classification layer is replaced as to fit the new task. Cross entropy loss uses the raw outputs of this layer (conform \texttt{PyTorch} documentation), together with one-hot encoded prediction targets. An early stop occurs after 10 epochs without improved loss on the validation set, and the model with the lowest loss is benchmarked on the testing set.

For OTS learning, the Adam optimizer \citep{kingma2014adam} is used with standard \texttt{PyTorch} parameters (\texttt{lr}=1e-3, $\beta_1=0.9$, $\beta_2=0.999$), and a batch size of 256. For FT, more is needed to achieve good convergence, especially for VT-based architectures. Used hyperparameters, etc. are partially inspired by \citeauthor{matsoukas2021time} (\citeyear{matsoukas2021time}) and \citeauthor{zhou2021convnets} (\citeyear{zhou2021convnets}). The Adam learning rate now starts off at 1e-4, and is reduced by a factor 10 after 3 epochs without improvement (i.e. patience=2); the batch size is reduced to 32; label smoothing of 0.1 is added to the cross entropy loss, and a dropout layer with p=0.2 is inserted before the classification layer. Finally, input images are augmented with random horizontal flips and rotations in a $\pm 10 ^\circ$ range.

\subsection{Hardware and software}
All experiments are conducted on a single compute node containing one 32 GB Nvidia V100 GPU. These nodes are constituents of the \textit{Peregrine high performance computing cluster}, belonging to the \textit{University of Groningen's Center for Information Technology}. The FT experiments take advantage of the V100's mixed precision acceleration. An exception is made for the type classification experiment, as this one is also used to compare OTS learning and FT in terms of time/accuracy trade-offs. Preliminary findings suggested that using mixed precision does not affect performance for the aforementioned experiments, however.

The \texttt{PyTorch} machine learning framework is used \citep{paszke2019pytorch}, and many pre-trained models are taken from its \texttt{Torchvision} library. Exceptions are \texttt{EfficientNetV2}, \texttt{Swin}, \texttt{DeiT} and \texttt{BeiT}, which are taken from the \texttt{Timm} library\footnote{\url{https://timm.fast.ai/}.}.

>>>>

% Experiments build upon Sabatelli et al.
Much of the experiments and analysis discussed below, build upon the work done by \citeauthor{sabatelli2018deep} (\citeyear{sabatelli2018deep}). This paper investigated TL properties of CNNs in the art classification domain. The main focus was on a comparison between CNN architectures pre-trained on ImageNet1K when they were either used as OTS feature extractors, or fine-tuned to the new task. It was suggested that fine-tuning resulted in substantially better performance than either training from scratch or using an OTS learning scheme.

@@

Similar to \citeauthor{sabatelli2018deep}, used models are pre-trained on ImageNet1K, and the \textit{Rijksmuseum Challenge} dataset \citep{mensink14icmr} is used for the target tasks. This dataset consists of a large collection of digitized artworks, together with xml-formatted metadata. The classification tasks extracted from it are: (1) \textit{Type classification}, where the model has to distinguish between classes as `painting', `sculpture', `drawing', etc.; (2) \textit{Material classification}, with classes as `paper', `porcelain', `silver', etc.; and finally, (3) \textit{Artist classification}, where the model has to predict who the creator is. Figure \ref{methods:eg} shows labeled examples for all three tasks.

@@

There are still many samples and classes left after doing the operations described above. Samples are also unevenly distributed among classes, with Gini-coefficients being as high as 0.98\footnote{The Gini-coefficient is a measure of balance, with 0 meaning all classes have the same number of samples, and 1 meaning all samples belong to a single class. A uniform distribution has coefficient $\frac{1}{3}$.}. To counteract this, only the 30 most occurring classes are selected, and a cap of 1000 instances per class is enforced by taking a random sample when this limit is exceeded. Table \ref{methods:datasets} shows the result of these balancing and subsampling operations. The situation prior is shown between brackets.

@@

All datasets are split up into 80\%, 10\%, 10\% training, validating, and testing sets. For robustness, five datasets are randomly generated per classification task, such that each experiment can be run five times. The average sample overlap between two of the five sets is shown in table \ref{methods:datasets} as well.

Besides the tasks described above, table \ref{methods:datasets} shows an experiment called \textit{Scaling}. The goal of this experiment is to see how the findings of this paper hold up as dataset sizes are gradually shrunk. It does this by taking the top 15 classes of type classification, and scaling it 4 times by a factor $\sqrt[4]{\frac{1}{10}} \approx 0.56$. The 5 datasets produced in this manner are then respectively 100\%, 56\%, 32\%, 18\%, and 10\% the size of the dataset shown in table \ref{methods:datasets}. Steps are taken to ensure that the distribution remains the same. Finally, also here 5 datasets are generated per scaling factor, leading to a total of 25 datasets. The same 80\%, 10\%, 10\% split is used for all.

@@

For all models, the linear classification layer is replaced to fit the new task. Cross-entropy loss uses the raw outputs of this layer (conform \texttt{PyTorch} documentation), together with one-hot encoded prediction targets. An early stop occurs after 10 epochs without improved loss on the validation set, and the model with the lowest loss is benchmarked on the testing set.

For OTS learning, the Adam optimizer \citep{kingma2014adam} is used with standard \texttt{PyTorch} parameters (\texttt{lr}=1e-3, $\beta_1=0.9$, $\beta_2=0.999$), and a batch size of 256. For FT, more is needed to achieve good convergence, especially for VT-based architectures. Used hyperparameters, etc. are partially inspired by \citeauthor{matsoukas2021time} (\citeyear{matsoukas2021time}) and \citeauthor{zhou2021convnets} (\citeyear{zhou2021convnets}). The Adam learning rate now starts off at 1e-4, and is reduced by a factor 10 after 3 epochs without improvement (i.e. patience=2); the batch size is reduced to 32; label smoothing of 0.1 is added to the cross-entropy loss, and a dropout layer with p=0.2 is inserted before the classification layer. Finally, input images are augmented with random horizontal flips and rotations in a $\pm 10 ^\circ$ range.

\subsection{Hardware and software}
All experiments are conducted on a single compute node containing one 32 GB Nvidia V100 GPU. These nodes are constituents of the \textit{Peregrine high-performance computing cluster}, belonging to the \textit{University of Groningen's Center for Information Technology}. The FT experiments take advantage of the V100's mixed precision acceleration. An exception is made for the type classification experiment, as this one is also used to compare OTS learning and FT in terms of time/accuracy trade-offs. Preliminary findings suggested that using mixed precision does not affect performance for the aforementioned experiments, however.

The \texttt{PyTorch} machine learning framework is used \citep{paszke2019pytorch}, and many pre-trained models are taken from its \texttt{Torchvision} library. Exceptions are \texttt{EfficientNetV2}, \texttt{Swin}, \texttt{DeiT}, and \texttt{BeiT}, which are taken from the \texttt{Timm} library\footnote{\url{https://timm.fast.ai/}.}.
